---
title: "Big Data Project -\n Prediction of departure delay flights"
author: "Yaara Diamant Karasik & Tal Goldberg"
date: "2022-08-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Flight delays are a significant societal problem as they impair airlines, transport companies, air traffic controllers, facility managers, and passengers. Studying flight data is an essential activity for every player involved in the air transportation system. Given the uncertainty of delays, passengers usually plan to travel many hours earlier for their appointments. They might have to increase their travel costs to arrive on time. Airlines suffer penalties, fines, and additional operation costs, such as crew and aircraft retention in airports  

```{r cars}
summary(cars)
```

## Research Aim

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


## Data

Our analysis is based on public flights dataset taken from "nycflights13" R package.This package contains information about all flights that departed from NYC at 2013. We used additional datasets from this package which provide extensive information about the flights. Thus we merged the 'planes' dataset which containes a construction information about each plane, by 'tailnum' variable column. 


## Exploratory data analysis (EDA)




## Inference



## Conclusion